---
title: "FUNNEL-GSEA R Package Vignette"
author: "Yun Zhang"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{FUNNEL-GSEA R Package Vignette}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---
```{r global_options, include=FALSE}
knitr::opts_chunk$set(warning=FALSE, message=FALSE)
```
                      
# Introduction

```{r package}
library(FUNNEL)
```

This package is built for ... In this vignette, I will introduce two ways of doing the FUNNEL test. One way is through a wrapper function with will take the raw cleaned data sets as inputs and run the whole testing procedure automatically. Secondly, I will guide the users to go through the testing framwork step-by-step, so that the users may have more flexibity and control over their GSEA tests. 

# One-step test

Firstly, let's load in the data.
```{r data}
data("sampleData")
```
The one-step test is contained in the function called `FUNNEL.GSEA`. It takes the raw and cleaned data as input and do the data formating and run the whole testing procedure automatically. For details of data formating, please see `help(sampleData)`. (It will take about 5~10 minutes to run.)
```{r, onestep, cache=TRUE}
wrapper <- FUNNEL.GSEA(X, time, geneset)
```

## Some post-analysis tools

* If the users are interested in some particular genes, they can gether them as a vector in 'genesOfInterest' and use the following function based on the one-step testing results.
```{r}
est.weights <- weightPerGene(wrapper$weight.list, genesOfInterest=geneset[[1]])
```

* Also, the users can use the previous results to plot the non-zero weights for a partucular gene set. For example, look at the estimated non-zero weights in the second gene set. 
```{r, weightplot, fig.width=6, fig.height=12}
plotWeight(wrapper$weight.list, geneset.index=2)
```

# Step-by-step analyses

# Session Info

```{r}
sessionInfo()
```

# The following is template

Vignettes are long form documentation commonly included in packages. Because they are part of the distribution of the package, they need to be as compact as possible. The `html_vignette` output type provides a custom style sheet (and tweaks some options) to ensure that the resulting html is as small as possible. The `html_vignette` format:

- Never uses retina figures
- Has a smaller default figure size
- Uses a custom CSS stylesheet instead of the default Twitter Bootstrap style

## Vignette Info

Note the various macros within the `vignette` section of the metadata block above. These are required in order to instruct R how to build the vignette. Note that you should change the `title` field and the `\VignetteIndexEntry` to match the title of your vignette.

## Styles

The `html_vignette` template includes a basic CSS theme. To override this theme you can specify your own CSS in the document metadata as follows:

    output: 
      rmarkdown::html_vignette:
        css: mystyles.css

## Figures

The figure sizes have been customised so that you can easily put two images side-by-side. 

```{r, fig.show='hold'}
plot(1:10)
plot(10:1)
```

You can enable figure captions by `fig_caption: yes` in YAML:

    output:
      rmarkdown::html_vignette:
        fig_caption: yes

Then you can use the chunk option `fig.cap = "Your figure caption."` in **knitr**.

## More Examples

You can write math expressions, e.g. $Y = X\beta + \epsilon$, footnotes^[A footnote here.], and tables, e.g. using `knitr::kable()`.

```{r, echo=FALSE, results='asis'}
knitr::kable(head(mtcars, 10))
```

Also a quote using `>`:

> "He who gives up [code] safety for [code] speed deserves neither."
([via](https://twitter.com/hadleywickham/status/504368538874703872))
